---
title: "week3_part2"
author: "Fiona Kate Morgan"
date: "2022-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1 - Synthetic diamonds hypothesis testing
A company produces synthetic diamonds that have a mean weight of 0.5 carat.
An experiment has been conducted to evaluate a new process for producing synthetic diamonds. Six diamonds have been generated by the new process, with recorded weights of 0.46, 0.61, 0.52, 0.48, 0.57 and 0.54 carat.
It is essential that the new process produces diamonds with a weight in excess of 0.50 carat.

Do the six diamond weight measurements present sufficient evidence to indicate that the mean weight of the diamonds produced by the process is in excess of 0.5 carat? Use R studio to test at the 1% significance level.

## Initialise the vector
```{r}
diamondsData <- c(0.46, 0.61, 0.52, 0.48, 0.57, 0.54)
```

## Define the null hypothesis
The hypothesis to be tested is:
$H_0: \mu >0.5$ and $H_1:\mu<=0.5$

## Decide on the required test
Because it is related to a mean I will use the t.test, and as the alternate hypothesis is 'not equal' it will be a single-sided test. 

```{r}
t.test(diamondsData, mu=0.5, alternative = "less")
```

## conclusion
From this output we can see that the p-value is 0.8773, so with 99% confidence level ($\alpha=0.01$)
The p-value is larger than the alpha so we would not reject $H_0$ based on this sample of data and therefore can conclude that it can be statistically possible that this sample of data comes from a population of diamonds that has a mean weight of >0.5 carat. 
