---
# This is the YAML header
title: "CS5702 Week3 Lab Notebook"
output:
  html_document: default
  pdf_document: default
author: Martin Shepperd
date: 01/10/2022
---

# Week 3 Worksheet 

Be aware that because this Worksheet contains some intentional (and hopefully no unintentional) coding errors you can't Knit it to html or pdf until these are fixed.  However, you can still save it as an RMD file in the usual way.  

## Worksheet Contents

0. [Worksheet: Introduction](#W3S0)
1. [Seminar: Visualising covid-19 data trends](#W3S1)
2. [Lab: Debugging and getting help](#W3S2)  
3. [Lab: User defined functions](#W3S3)    
4. [Exercise answers](#W3A) 

##0. Worksheet Introduction {#W3S0}  

<span style="color: darkorange;">**Pre-requisites**</span>

You should:  

1. have worked through the Week 2 Worksheet 
2. be familiar (listened to/read) the Week 3 Lecture "Engineering or Hacking"
3. be able to write, edit, save and re-open your own RMarkdown files  
4. be able to load external files (of various formats such as csv) into a data frame

This lab worksheet is organised as an RMarkdown file.  You can **read** it.  You can **run** the embedded R and you can **add** your own R.  I suggest you save it as another file so, if necessary, you can revert to the original.  

Remember, you are encouraged to explore and experiment (although try to do this systematically).  Change something and see what happens!

### 0.1 Packages and libraries

Up until now we have mainly dealt with Base R, principally for reasons of simplicity.  This is often a sufficient set of functions to perform straightforward data and statistical analysis.  However, there is a rich world of **additional** functions that are freely available.  

We can **extend** Base R by means of <span style="color: darkorange;">**packages</span> (collections of extra functions and data sets) developed by the R community and validated by [CRAN](https://cran.r-project.org/), which hosts most packages.  Presently there are about 18,000 extra packages covering everything from graphics to machine learning.  

In order to install a new package you need to fetch it from CRAN and copy onto a local storage device as part of your R library e.g., your hard disk.  You do this **ONCE**.  Thereafter, if you need to use it you must copy it into memory each R session.

```{r}
# Install a library
# Only do this once!
# Uncomment the next statement if you wish to execute it.

# The packages we need for this tutorial are:
# (1) tidyverse - a useful collection of functions embedded in
# various packages including {ggplot2}
# (2) devtools - to enable us to load packages from GitHub rather than CRAN
# (3) tidycovid isn't on CRAN so we need to install from GitHub

# To do this you need to uncomment the first time
# install.packages("tidyverse")
# install.packages("devtools")
# 
# NB You must install devtools *before* tidycovid19

# Then *every* new R session
library(tidyverse)
library(devtools)
library(tidycovid19)


install_github("joachim-gassen/tidycovid19")



# To check what packages you already have in your library
# installed.packages()
# NB this can be a very long list after a while so one option is to save it as an R
# data object e.g.,
PkgList <- installed.packages()
```

**Exercise 0.1:** What type is `PkgList`?  How might you conveniently explore it?  [HINT: View]

```{r}
View(PkgList)
```


On rare occasions a package may be on GitHUb rather than CRAN e.g., {tidycovid19} in which case you need the package {devtools} which is on CRAN(!) and then use the function `install_github("joachim-gassen/tidycovid19")` which gives the Git repo name `joachim-gassen` and the package name `tidycovid19`.

There can sometimes be complex dependencies between packages and so installing one package may lead to others being installed.  Similarly, when you load from your R library.  For this reason it's better to avoid repeatedly re-installing packages.


## 1. Seminar - Visualising covid-19 data trends {#W3S1}

This example shows how we can fetch and visualise covid-19 data from Johns Hopkins University via GitHub.

### 1.1 Initialisation

We need some packages over and above base R.  Since we may not be sure whether they are already installed we test for their presence.  Most packages come from CRAN and are easy to install using `install.packages()` but the package `{tidycovid19}` is on GitHub (joachim-gassen) so we also need `{devtools}` in order to install packages that aren't on CRAN.

This R code may appear daunting but don't worry.  We will revisit it in detail in Week 3.  For the time being see it as a way to install and load necessary extra functionality beyond Base R.

```{r messages=F}

```

Download the data (cached on GitHub rather than directly from the Johns Hopkins University).  This is live data updated within the last 24 hours.  

```{r}
#Download the data into a data frame called cv.df using the 
#download_jhu_csse_covid19_data() function from the {tidycovid19} package.
#
cv.df <- download_jhu_csse_covid19_data(cached = TRUE)
```

**Exercise 1.1:** The dataframe which comprises all international covid-19 data recorded by Johns Hopkins since January 22, 2020 has ~191,000 observations (see the Environment Pane).  Is this large?  How much larger can R handle?


### 1.2 Explore the data

Let's focus on the UK  and then "eyeball" the data again.

```{r}
# select only the UK data
cv.uk.df <- subset(cv.df, iso3c=="GBR")

head(cv.uk.df)   # Defaults to first 6 rows/observations
tail(cv.uk.df)   # Defaults to last 6 rows/observations
```

### 1.3 Show trends

(i) Mortality rate

```{r messages=F}
# Compute new deaths as the data shows cumulative deaths
cv.uk.df$new.d[2:nrow(cv.uk.df)] <- tail(cv.uk.df$deaths, -1) - head(cv.uk.df$deaths, -1)
cv.uk.df$new.d[1] <- 0     # Add zero for first row 

# Compute new infections
cv.uk.df$new.i[2:nrow(cv.uk.df)] <- tail(cv.uk.df$confirmed, -1) - head(cv.uk.df$confirmed, -1)
cv.uk.df$new.i[1] <- 0     # Add zero for first row 

head(cv.uk.df)
```

We can produce a plot of daily additional deaths using the {ggplot} package which is an extremely powerful and flexible set of functions for producing extremely high quality graphics e.g. NYT, Guardian and the BBC.  We also save the plots using the `ggsave()` function which is also part of the {ggplot} package.

```{r}
# NB a small span value (<1) makes the loess smoother more wiggly!
ggplot(data = cv.uk.df, aes(x = date, y = new.d)) +
  geom_line(color = "skyblue", size = 0.6) +
  ylim(0,1200) +
  stat_smooth(color = "darkorange", fill = "darkorange", method = "loess", span = 0.2) +
  ggtitle("Daily additional deaths in the UK due to covid-19") +
  xlab("Date") + ylab("Daily new deaths")
ggsave("cv19_UK_deathrate.png")
```

(ii) Infection rate


```{r}
ggplot(data = cv.uk.df, aes(x = date, y = new.i)) +
  geom_line(color = "skyblue", size = 0.6) +
  scale_y_continuous(trans = "log10") +
  stat_smooth(color = "darkorange", fill = "darkorange", method = "loess") +
  ggtitle("Daily new infections in the UK from covid-19") +
  xlab("Date") + ylab("Daily new infections") 
ggsave("cv19_UK_infectionrate.png")
```


**Exercise 1.2:** Note that we use a log scale for the y-axis ie daily new infection rate.  Why?
Becaue we're showing a change that is much smaller than the bulk of the data.

To better visualise the trends (i.e., over time) we use the **loess** (locally estimated scatterplot smoothing) smoother.  It is designed to detect trends in the presence of noisy data when the shape of the trend is unknown thus it is a robust (non-parametric) method.  


**Exercise 1.3:** If you look carefully at the data there is a clear cycle within the overall trend.  What is it and why?  How should we deal with it?  What does the light orange shaded region around the smoothed trend line mean?  Why does it vary in breadth?

The cyclic data will be due to more deaths being reported on certain days of the week. I assume this is possibly due to more staff, specifically admin staff who process the death reports, more likely working weekdays.
It would be interesting to dive into this deeper, although it would be far more interesting if the deaths were actually more likely to be happening on particualar days of the week. In this case, if my assumption is correct, as I am looking more generally at the change over a longer period of time, the specific weekly cyclic nature of the data is not of interest.

The light orange is the confidence interval around the smoothed line. 

**Exercise 1.4:** What was the greatest number of new infections in one day in the UK?  HINT there is a built in function `max()` and the you will need to examine the vector `new.i` which is part of the `cv.uk.df` dataframe so you will need the `$` operator.

```{r}
max(cv.uk.df$new.i)
```
The maximum number of new infections in a single day in the UK was 848,169

**Exercise 1.5:**  The plots cover a long period of time and are not easy to read.  Suppose we want to focus on 2022?  To do this we can use a subset based on date

```{r}
# Create a new data frame for 2022 only (from the UK).

cv.uk2022.df <- cv.uk.df[cv.uk.df$date>="2022-01-01", ]
```

Now repeat the time series using only 2022 data.  Experiment with other subsets.

```{r}
ggplot(data = cv.uk2022.df, aes(x = date, y = new.i)) +
  geom_line(color = "skyblue", size = 0.6) +
  scale_y_continuous(trans = "log10") +
  stat_smooth(color = "darkorange", fill = "darkorange", method = "loess") +
  ggtitle("Daily new infections in the UK from covid-19 - 2022") +
  xlab("Date") + ylab("Daily new infections") 
ggsave("cv19_UK2022_infectionrate.png")
```
```{r}
# A new data frame for 2021 only (from the UK).

cv.uk2021.df <- cv.uk.df[(cv.uk.df$date>="2021-01-01" & cv.uk.df$date<"2022-01-01"), ]
head(cv.uk2021.df)
```


```{r}
ggplot(data = cv.uk2021.df, aes(x = date, y = new.i)) +
  geom_line(color = "skyblue", size = 0.6) +
  scale_y_continuous(trans = "log10") +
  stat_smooth(color = "darkorange", fill = "darkorange", method = "loess") +
  ggtitle("Daily new infections in the UK from covid-19 - 2021") +
  xlab("Date") + ylab("Daily new infections") 
ggsave("cv19_UK2021_infectionrate.png")
```


**Extension Exercise 1.6:** Edit the R code to produce a similar trend analysis for another country of your choice.  Note that the data set uses 3 character country codes e.g., SWE, USA see [wikipedia](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) for a complete list.  HINT: you will need to change the subset command and perhaps save to a new dataframe and then make sure you turn the cummulative counts into new counts.

```{r}
cv.de.df <- subset(cv.df, iso3c=="DEU")

# Compute new deaths as the data shows cumulative deaths
cv.de.df$new.d[2:nrow(cv.de.df)] <- tail(cv.de.df$deaths, -1) - head(cv.de.df$deaths, -1)
cv.de.df$new.d[1] <- 0     # Add zero for first row 

# Compute new infections
cv.de.df$new.i[2:nrow(cv.de.df)] <- tail(cv.de.df$confirmed, -1) - head(cv.de.df$confirmed, -1)
cv.de.df$new.i[1] <- 0     # Add zero for first row 

head(cv.de.df)
```

```{r}
cv.de2022.df <- cv.de.df[cv.de.df$date>="2022-01-01", ]
head(cv.de2022.df)

ggplot(data = cv.de2022.df, aes(x = date, y = new.i)) +
  geom_line(color = "skyblue", size = 0.6) +
  scale_y_continuous(trans = "log10") +
  stat_smooth(color = "darkorange", fill = "darkorange", method = "loess") +
  ggtitle("Daily new infections in Germany from covid-19 - 2022") +
  xlab("Date") + ylab("Daily new infections") 
ggsave("cv19_DE2022_infectionrate.png")
```



**Exercise 1.7:** Is `new.i` in the cv.uk.df dataframe an integer?  

```{r}
typeof(cv.uk.df$new.i)
```
No - it's a double. 

**Exercise 1.8:** How many people recovered in the UK on the 236th day of data collection.  Write an R statement to answer. 

```{r}
print(cv.uk.df[236,"recovered"])
```
there were 2,148 recoveries in the UK on the 236th day of collection.


## 2. Lab - Debugging and getting help {W3S2}

1. find/read the relevant [cheatsheet](https://rstudio.com/resources/cheatsheets/)  
2. perspiration e.g., see this [five step approach](https://medium.com/learn-love-code/stuck-on-a-coding-problem-here-are-5-steps-to-solve-it-8be04c4b4f19)
3. talk it over with a fellow student
4. module **FAQs** on Brightspace (this will grow over time)  
5. [Stack overflow](https://stackoverflow.com/)  
6. asking a member of the course team  

For *more* suggestions visit the subsection 0.2 ["vi) Learn how to get help"](https://bookdown.org/martin_shepperd/ModernDataBook/W0-Prep.html#Help) in the Modern Data book.  

**Exercise 2.1:** The following R code chunk contains a bug such that the R interpreter can't parse it.  What's wrong?  Can you fix it?

```{r}
# This code chunk contains an error
var1 <- 5
var2 <- 6
# var3 <- (var1 * var2)/(varI * 100) - the error was with the varI variable, it doesn't exist so I changed it to var1
var3 <- (var1 * var2)/(var1 * 100)
```

**Exercise 2.2:** This R chunk has a logic error.  Can you detect it and then make the necessary correction?  It might help to look at the values of the variables in the Environment Pane.

```{r}
# This code chunk contains an error
var4 <- 1
var5 <- 2
var6 <- (var4 + var5)/(var4 * var5)
if(var6 > var4) {
  print("var6 is larger than var4")
} else if(var6 < var4) {
  print("var6 is smaller than var4")
} else {
  print ("va6 is the same as var4")
}
```
The greater than and less than signs were the wrong way around.

In addition, the above code doesn't cover all cases which makes it a bit 'ugly'. Can you fix the problem? 
I changed it to a multiple if else statement.

**Exercise 2.3:** The following R code chunk contains a bug such that yet again the R interpreter can't parse it.  What's wrong?  Can you fix it?

The 0 was an upcapped o

```{r}
# Here we are trying to create a numeric vector using
# the combine c() function.  But there's an error.
Vector1 <- c(3, 2, 0, 2, 2, 6)
```

**Exercise 2.4:** The following R code chunk contains a bug such that yet again the R interpreter can't parse it.  What's wrong?  Can you fix it?
x was written as a string "1" so I removed the speechmarks to make it an integer like the rest of the vector.

```{r}
x <- 1
Vector2 <- c(3, 2, 0, 2, 2, 6)
Vector2 <- c(Vector2, x)
sum(Vector2)
```


## 3. User defined functions {#W3S3} 

> One of the best ways to improve your reach as a data scientist is to write functions. Functions allow you to automate common tasks in a more powerful and general way than copy-and-pasting.  
> --- Grolemund and Wickham (2018) 

User-defined function *benefits*:  

- abstract (hide) details to help understandability e.g., `checkInput()`   
- place related code in one place    
- reuse  
- maintainability i.e., only need to make changes in *one* place  


Parts of a function:  

1. **Function Name**: how we call the function e.g., `head()`.  Try to use descriptive names.
2. **Argument(s)**: when a function is called, we pass values to match the (0+) arguments.
3. **Function Body**: contains a set of statements that defines what the function does.  (For in-built functions these are hidden but for user defined-functions we need to provide them.)  
4. **Return Value**: the result after being called i.e., what's given back to the calling code.


We can DEFINE a function called *halve* by declaring it.  Afterwards you can see it as a new R object in the Environment Pane.

```{r}
halve <- function(n)
# This function takes a number and divides it by two
{
  functionResult <- n/2  
  return(functionResult) 
}
```

We USE the function by calling it *after* it has been successfully declared.

``` {r}
halve(-8.3)
```

**Exercise 3.1:** Create a new function named `third()` that returns a third of its input argument n.  Check it works by testing it with various input values, e.g., `third(7)`.

```{r}
third <- function(n)
{
  functionResult <- n/3  
  return(functionResult) 
}

third(15)
third(300)
```



**Exercise 3.2:** Generalise the function to `fraction()` so that it now has two arguments where the first is the numerator and the second is the denominator.  Declare the function and again test it with different input values, e.g., `fraction(7, 10)`.

```{r}
fraction <- function(n, d)
{
  functionResult <- n/d  
  return(functionResult) 
}

fraction(1,2)
fraction(5,100)
```


**Exercise 3.3:** Write a new function `notNegative()` that takes a number as an argument and return TRUE if it's zero or more (i.e., not negative) and FALSE otherwise.  You will need an `if` statement that tests if a condition is true or not, e.g., `if (x<y) ... else ...`.  For more details use the help e.g., enter "if' into the search box of the Help Pane in RStudio.

```{r}
notNegative <- function(n) 
{
  if (!is.numeric(n)) {
    return(NULL)
  }
  
  if (n < 0) {
    return (FALSE)
  } else {
    return (TRUE)
  }
}

notNegative(-3)
notNegative(3)
notNegative(0)
notNegative("2")
```



**Extension exercise 3.4:** What happens to your function if you give a value that's not numeric e.g., a character?  How might you protect your function against this eventuality?  This is sometimes known as defensive programming.  HINT: There's a built-in function `is.numeric()`. You can discover more about it using the help, e.g., type `?is.numeric`.


## 4. Exercise Answers{#W3A}

1.1: Although ~120K observations might seem large, in reality this only occopies 2.6Mb which is <0.1% of the capacity of R on a fairly standard laptop or PC.

1.2: Given the wide range of values for daily infections a log10 scale makes the plot easier to view, particularly for the smaller values. 

1.3: There is a clear weekly cycle (or we can say the periodicity is 7-days).  This is true of many countries.  Why do you think this might be?  

The orange band  shows the 95% confidence limit since there is an element of uncertainty as to exactly where the trend line should be.  The broader the confidence limit band (shaded pale orange) the less confident we are about the exact location of the trend.  Where the confidence limit potentially goes negative (which would be meaningless) we do not plot it.  This principally occurs for the daily new death rate trend since many values are (mercifully) close to zero. 

There are many ways to compute a confidence interval.  A common and robust method is known as the bootstrap.  See this [short article](https://acclab.github.io/bootstrap-confidence-intervals.html) for an easy introduction.

1.4: `max(cv.uk.df$new.i)`  

1.6: Good luck!  

1.7: `new.i` is not an integer?  One way to find out is to use the built in function `is.integer()`.
```{r}
is.integer(cv.uk.df$new.i)
```

If you wanted `new.i` to be an integer you could need to code something like
```{r}
cv.uk.df$new.i <- as.integer(cv.uk.df$new.i)
```
or make an assignment like `cv.uk.df$new.i <- 10L` where L is short for Long which is a long story!  

1.8: `cv.uk.df$recovered[236]`  Remember you need to use the $ operator to reference a particular vector (variable) in the dataframe cv.uk.df.


2.1: Although superficially the code looks ok, `var1` has been mistyped as `varI` on line 49.  Variable names need to be exact including the same case.

2.2: The logical problem lies in the two if statements.  The relational operators (< and >) are the wrong way round, i.e., they're not consistent with the message from `print()`.  Making errors with `if` conditions is all too easy.  Be on the look out for this type of mistake.  

2.3: As per 3.1 superficially the code looks ok but if you inspect the arguments to c() carefully you will see that insetad of a zero (0) an upper case O has been typed.  A hint is the editor has coloured it black rather than blue which RSTudio uses for numeric literals, black indicates it may be a variable name.  Since it's not in quotes R assumes it's the name of an object (variable) called `O` and since this has been declared it doesn't know what to do and gives up.  A solution is to replace `O` with 0.

2.4: The problem is to do with type.  If you look carefully you can see that x is a character type because the number assigned to it is in quotation marks.  Therefore when we combine it with the numeric vector Vector2 this **coerces** the type to a character vector.  Remember every element of a vector must be the same type.  One the type has changed from numeric to character it's no longer meaningful to apply the arithmetic function `sum()` to it.

3.1: This change should be straightforward.  Don't forget to edit the comment.  Obsolete and redundant comments are extremely harmful to the goal of understandable code.  

```{r}
third <- function(n)
# This function takes a number and divides it by three
{
  functionResult <- n/3  
  return(functionResult) 
}
```

3.2: Here is a possible solution.  Note that there are now two arguments to this function.  The comment also been added to explain the purpose of each argument particularly because `n` and `m` are quite gnomic. 

```{r}
fraction <- function(n,m)
# This function takes a number n (numerator) and divides it by m (denominator)
{
  functionResult <- n/m  
  return(functionResult) 
}
```

3.3: This function requires a logical test using an `if` statement. Again this function might seem trivial but it's good to add a comment to properly document it.  Remember the direction we're testing, i.e., *not* negative so zero or more should return TRUE.  When you test your function, e.g., `notNegative(-3)` remember to cover positive and negative cases and it's good practice to also consider the boundary case (which is zero).

```{r}
notNegative <- function(n){
# This function checks if n is negative (returns FALSE) or not (returns TRUE)
  if (n<0) return(FALSE) 
  else return(TRUE)
  }
```


3.4: Good luck!  

This version tests if the argument is numeric using the built-in function `is.numeric()`.

Without changing the return data type from logical which would turn the function into something very different I assume that non-numeric types cannot be considered as not negative so the function returns FALSE if either n isn't numeric or it isn't non-negative.  

This requires a nested `if ... else`.

```{r}
notNegative <- function(n)  
{
# This function checks if n isn't numeric and returns FALSE 
# else if n is numeric then 
# tests if it is not negative ie zero or more (returns TRUE) else (returns FALSE)

    if (!is.numeric(n))    # The ! is a logical not ie not numeric
      return(FALSE)
    else
      if (n<0) return(FALSE)
      else
        return(TRUE)
  }
```

A more elegant solution removes the need for negation.

```{r}
notNegative <- function(n)
{
  # This function checks if n isn't numeric and returns FALSE
  # else if n is numeric then
  # tests if it is not negative ie zero or more (returns TRUE) else (returns FALSE)
  
  if (is.numeric(n))
    if (n < 0)
      return(FALSE)
    else
      return(TRUE)
  else
    return(FALSE)
}
```

