---
title: "weather project"
output: html_document
date: "2022-09-19"
---

No idea what this is
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Import ggplot2 library
```{r}
library(ggplot2)
```

2) Assign weather data csv to 'weather' variable 
```{r}
weather <- read.csv("weather.csv")
```

```{r}
summary(weather)
```

caluclate mode function
```{r}
findmodes <- function(a) {
  u_val <- unique(a)
  tab <- tabulate(match(a, u_val))
  u_val[tab == max(tab)]
}
```


3) mean, mode and median of Temp_C column
```{r}
mean <- mean(weather$Temp_C)
mode <- findmodes(weather$Temp_C)
median <- median(weather$Temp_C)

mean
mode
median
```

4) frequency table for 'Weather' column
```{r}
table(weather$Weather)
```

5) Scatterplot of Temperature and Humidity
```{r}
plot(weather$Temp_C, weather$Rel.Hum_., main="Temperature & Humidity", xlab="temp", ylab="humidity")
```
Between the temperatures of 5 and 24 humdity can generally be any value, although is most likely to be above 40. At about 24 degrees the humidity is less likely to be above 70. 


6) Same graph plotted using ggplot2
```{r}
ggplot(weather, aes(x=Temp_C, y=Rel.Hum_.)) + geom_point()
```

7) Histogram showing tempratures
```{r}
hist(weather$Temp_C, xlab="temp", col="pink")
```

